LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-10-06 main.tex
Add FIXME comments from J. Comp. Sci. peer review

\begin{abstract}
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool. LCSs are most repelling, attracting and shearing material lines that form the centerpieces of observed tracer patterns in two-dimensional unsteady dynamical systems. LCS Tool implements the latest geodesic theory of LCSs for two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations. After a review of the underlying theory, we explain the steps and numerical methods used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples.
\end{abstract}

@@

Lagrangian Coherent Structures (LCSs) are evolving organizing centers of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.

@@

Here we present a simple algorithmic introduction to geodesic LCS detection without details on the the underlying mathematics.
% FIXME Replace GitHub link with http://runmycode.org/
We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is publicly available at: https://github.com/jeixav/LCS-Tool}.
This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present here have been programmed into demonstration scripts.

The list of publicly available software for LCS detection further
includes the following packages:
\begin{itemize}
\item \emph{MANGEN}\citep{lekien03:_time} calculates FTLE and advects material curves in two dimensional velocity fields. It also includes a graphical user interface and uses MPI for parallel calculations.
% FIXME Cannot cite LCS MATLAB Kit without a journal reference
% \item \emph{LCS MATLAB Kit}\citep{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\citep{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic  and dataset based velocity definitions.
% FIXME Cannot cite FlowVC without a journal reference
% \item \emph{FlowVC}\citep{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
% FIXME FlowTK
% https://github.com/FlowPhysics/FlowTK
% doi:10.1007%2F978-3-319-04099-8_13
\end{itemize}
These packages primarily target the automated generation of FTLE plots to aid the visual assessment of hyperbolic LCSs in the flow. The purpose of LCS Tool and this article is to provide an all-purpose numerical engine for the geodesic extraction of LCSs as parametrized material curves in unsteady two-dimensional flows, and extend the scope of such extraction to elliptic LCSs.

@@

@@

The Cartesian grid of initial conditions mentioned in \cref{t:Cauchy-Green algorithm} is rectangular, with user-defined vertical and horizontal ranges and resolutions. The optimal resolution may be determined by a successive doubling of the initial resolution until convergence of the extracted LCSs is observed visually. If the domain of interest comprises only a few expected LCSs, e.g., one vortex, then a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, a higher resolution must be chosen.

@@

To avoid excessive computational times in MATLAB, we vectorize \cref{eq:velocity}, i.e., combine its right hand side evaluated over each initial point into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ODEs. We then use MATLAB's \lstinline!ode45! function to perform trajectory integration from all grid points simultaneously. This calculation typically takes five to ten minutes with default error tolerances. A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

@@

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
\caption{Example of a Poincare map obtained for an elliptic LCS. Circle markers indicate closed orbit positions. The filled circle indicates the outermost fixed point of the Poincare map, marking the intersection of a coherent Lagrangian cortex boundary with the Poincare section.}
\label{f:Poincare return map}
\end{figure}

@@

@@

\begin{lstlisting}[caption={Double gyre derivative function corresponding to \cref{eq:double gyre derivative equations}.},label=l:double gyre derivative]
function derivative_ = derivative(t,x,epsilon,amplitude,omega)

@@

\lstinputlisting[
        caption={LCS Tool commands for double gyre hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m!},
        label=l:double gyre lambda strain stretch LCS]
        {listings/double_gyre/hyperbolic_lcs_subset.m}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_hyperbolic_lcs}
\caption{LCSs in the double gyre flow. Resolution $[500\times250]$. $\lambda = 1.00$ and $1.04$ for the left and the right gyre. $\lambda \in [0.93,1.07]$, $\Delta\lambda = 0.01$.
(Top) Elliptic (green) and repelling (red) LCSs with the FTLE shown in the background. (Bottom) Elliptic and attracting (blue) LCSs.}
\label{f:double gyre lambda hyperbolic LCS}
\end{figure}

@@

The integration time is $T=4L_x/U$, a multiple of a the eddy turnover time in the flow. \Cref{l:Bickley jet elliptic hyperbolic LCS subset} shows the LCS Tool commands in which the chaotically perturbed velocity is defined (\crefrange{ll:Bickley jet velocity definition start}{ll:Bickley jet velocity definition end}), periodic boundary conditions are imposed in the x-direction (\cref{ll:Bickley jet periodicBc}), and five Poincare sections are defined where we expect coherent vortices (\crefrange{ll:Bickley jet Poincare start}{ll:Bickley jet Poincare end}).
$\lambda$-values for closed orbit detection are varied over the range $[0.80,1.20]$ with a step of $0.01$.

@@

\lstinputlisting[caption={LCS Tool commands for Bickley jet elliptic and hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},
                 label=l:Bickley jet elliptic hyperbolic LCS subset]{listings/bickley_jet/elliptic_hyperbolic_lcs_subset.m}

>>>>

\begin{abstract}
% FIXME Reviewer #5, comment #1: Abstract and introduction the objective and contribution of the paper are not clear; several points need to be clarified further.
% FIXME Reviewer #5, comment #2: For readers to quickly catch the contribution in this work, it would be better to highlight major difficulties and challenges, and original achievements to overcome them, in a clearer way in abstract and introduction.
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool. LCSs are most repelling, attracting and shearing material lines that form the centerpieces of observed tracer patterns in two-dimensional unsteady dynamical systems. LCS Tool implements the latest geodesic theory of LCSs for two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations. After a review of the underlying theory, we explain the steps and numerical methods used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples.
\end{abstract}

@@

% FIXME Reviewer #5, comment #5: Please provide information regarding practical applications of this algorithmic software tool.
Lagrangian Coherent Structures (LCSs) are evolving organizing centers of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.

@@

Here we present a simple algorithmic introduction to geodesic LCS detection without details on the the underlying mathematics.
% FIXME JCompSci editor: Replace GitHub link with http://runmycode.org/
% FIXME Reviewer #3, comment #4: Add documentation within the github repository to show some reference to the APIs and how users may start to run a his/her own application.
% FIXME Reviewer #4: Please make sure the code is downloadable once the paper is published
We then review the implementation of this approach in a computational engine called LCS Tool\footnote{LCS Tool is publicly available at: https://github.com/jeixav/LCS-Tool}.
This engine is a library of MATLAB functions that extract LCSs from two-dimensional unsteady flows. The examples we present here have been programmed into demonstration scripts.

% FIXME Reviewer #3, comment #3: The comparison with other publicly available LCS detection software should include more details. It contains a very high-level summary of LCS tool being an "all-purpose numerical engine". That part should be revisited and become more concrete.
The list of publicly available software for LCS detection further
includes the following packages:
\begin{itemize}
\item \emph{MANGEN}\citep{lekien03:_time} calculates FTLE and advects material curves in two dimensional velocity fields. It also includes a graphical user interface and uses MPI for parallel calculations.
% FIXME Cannot cite LCS MATLAB Kit without a journal reference
% \item \emph{LCS MATLAB Kit}\citep{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\citep{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic  and dataset based velocity definitions.
% FIXME Cannot cite FlowVC without a journal reference
% \item \emph{FlowVC}\citep{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
% FIXME FlowTK
% https://github.com/FlowPhysics/FlowTK
% doi:10.1007%2F978-3-319-04099-8_13
\end{itemize}
These packages primarily target the automated generation of FTLE plots to aid the visual assessment of hyperbolic LCSs in the flow. The purpose of LCS Tool and this article is to provide an all-purpose numerical engine for the geodesic extraction of LCSs as parametrized material curves in unsteady two-dimensional flows, and extend the scope of such extraction to elliptic LCSs.

@@

% FIXME Reviewer #3, comment #1: Between the three applications, there should be a description about linking them in terms of generic APIs (application programming interface) or common code style. It is good to have code in the paper but it doesn't seem straightforward for a reader to find out how this simplifies the effort from coding from scratch (instead of using the platform). That part should be highlighted as a key point for judging the merit to the community. So, it would be very useful to pinpoint the generics with a flowchart (e.g. it always starts from input parameters, velocity definition and then LCS parameters, and how you finally call only 1-2 functions for the core solving routine), and your Matlab code there can further demonstrate the paradigm.
% FIXME Reviewer #5, comment #6: The functionality and taxonomy of the proposed LCS tool should be introduced for an overview understanding of the tool.

@@

The Cartesian grid of initial conditions mentioned in \cref{t:Cauchy-Green algorithm} is rectangular, with user-defined vertical and horizontal ranges and resolutions. The optimal resolution may be determined by a successive doubling of the initial resolution until convergence of the extracted LCSs is observed visually.
% FIXME Reviewer #2, comment #1: "500 grid point" → "500 grid points"
If the domain of interest comprises only a few expected LCSs, e.g., one vortex, then a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, a higher resolution must be chosen.

@@

To avoid excessive computational times in MATLAB, we vectorize \cref{eq:velocity}, i.e., combine its right hand side evaluated over each initial point into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ODEs. We then use MATLAB's \lstinline!ode45! function to perform trajectory integration from all grid points simultaneously. This calculation typically takes five to ten minutes with default error tolerances.
% FIXME Reviewer #2, comment #3: Add memory use data based on, for example, computations for \cref{f:double gyre lambda LCS convergence}.
A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

@@

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{graphics/double_gyre/poincare_return_map}
\end{center}
% FIXME Reviewer #2, comment #5: Indicate what are "s" and "P(s)"
\caption{Example of a Poincare map obtained for an elliptic LCS. Circle markers indicate closed orbit positions. The filled circle indicates the outermost fixed point of the Poincare map, marking the intersection of a coherent Lagrangian cortex boundary with the Poincare section.}
\label{f:Poincare return map}
\end{figure}

@@

% FIXME Reviewer #3, comment #2: "As for a platform, efficiency is also a matter in consideration. Performance degradation is not expected as a cost of generalized computational paradigm. That part should be at least mentioned along with the applications to make the article technically more sound."

@@

% FIXME Reviewer #2, comment #8: Check all code listings for awkward splits across pages and line-splits (e.g. semicolon on own line and array indices)
% FIXME Reviewer #5, comment #3: Format manuscript carefully with regards to code listings
\begin{lstlisting}[caption={Double gyre derivative function corresponding to \cref{eq:double gyre derivative equations}.},label=l:double gyre derivative]
function derivative_ = derivative(t,x,epsilon,amplitude,omega)

@@

% FIXME Reviewer #5, comment #4: There is too much source code in the main content such as cref{l:double gyre lambda strain stretch LCS,l:Bickley jet elliptic hyperbolic LCS subset,l:ocean data elliptic hyperbolic lcs}. They could be arranged as an appendix at the end.
\lstinputlisting[
        caption={LCS Tool commands for double gyre hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m!},
        label=l:double gyre lambda strain stretch LCS]
        {listings/double_gyre/hyperbolic_lcs_subset.m}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{graphics/double_gyre/lambda_hyperbolic_lcs}
% FIXME Reviewer #2, comment #6: Explain meaning of white dots ($\lambda$ maxima) and, if possible improve their appearance
\caption{LCSs in the double gyre flow. Resolution $[500\times250]$. $\lambda = 1.00$ and $1.04$ for the left and the right gyre. $\lambda \in [0.93,1.07]$, $\Delta\lambda = 0.01$.
(Top) Elliptic (green) and repelling (red) LCSs with the FTLE shown in the background. (Bottom) Elliptic and attracting (blue) LCSs.}
\label{f:double gyre lambda hyperbolic LCS}
\end{figure}

@@

% FIXME Reviewer #2, comment #7: Remove "a" in "multiple of a the eddy turnover", and sentence reads awkwardly
The integration time is $T=4L_x/U$, a multiple of a the eddy turnover time in the flow. \Cref{l:Bickley jet elliptic hyperbolic LCS subset} shows the LCS Tool commands in which the chaotically perturbed velocity is defined (\crefrange{ll:Bickley jet velocity definition start}{ll:Bickley jet velocity definition end}), periodic boundary conditions are imposed in the x-direction (\cref{ll:Bickley jet periodicBc}), and five Poincare sections are defined where we expect coherent vortices (\crefrange{ll:Bickley jet Poincare start}{ll:Bickley jet Poincare end}).
$\lambda$-values for closed orbit detection are varied over the range $[0.80,1.20]$ with a step of $0.01$.

@@

\lstinputlisting[
        caption={LCS Tool commands for Bickley jet elliptic and hyperbolic LCSs. Subset of the LCS Tool script \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},
        label=l:Bickley jet elliptic hyperbolic LCS subset]
        {listings/bickley_jet/elliptic_hyperbolic_lcs_subset.m}
